---
date: '10/23/2025'
title: Estimating harvest levels to reduce feral deer populations in Tasmania
categories: rstats
published: true
---

Feral deer populations in Tasmania have become a contentious political issue, balancing conservation priorities against harvest management. A [recent ABC news report](https://www.abc.net.au/news/2025-10-02/tas-new-feral-deer-survey-shows-population-boom/105843366) highlights the population boom and the challenges this presents.

In this post, I'll work through the mathematics of population growth and harvest management to understand what level of harvest would be needed to reduce deer populations back to historical levels.

## Population growth calculations

Let's start with the data: Tasmania's deer population has grown from 50,000 to 70,000 individuals over 5 years. We can calculate the annual finite rate of increase (λ) and convert it to an instantaneous rate of increase (r).

### Finite rate of increase (λ)

The finite rate of increase over a time period is given by:

$$ \lambda = \left(\frac{N_t}{N_0}\right)^{1/t} $$

Where:

- $N_0 = 50,000$ (initial population)
- $N_t = 70,000$ (population after 5 years)
- $t = 5$ years

Let's calculate this in R:

```{r}
# Initial parameters
N0 <- 50000  # Initial population
Nt <- 70000  # Population after 5 years
t <- 5       # Time period in years

# Calculate finite rate of increase
lambda <- (Nt / N0)^(1/t)
lambda
```

This means the population increases by about **`r round((lambda - 1) * 100, 2)`%** per year on a finite (discrete) basis.

### Instantaneous rate of increase (r)

We can convert λ to the instantaneous rate of increase using:

$$ r = \ln(\lambda) $$

```{r}
# Calculate instantaneous rate of increase
r <- log(lambda)
r
```

This corresponds to an instantaneous growth rate of **`r round(r * 100, 2)`%** per year, assuming continuous exponential growth.

## Required harvest rate

To reduce the deer population from 70,000 back to 50,000 over 5 years, we need to calculate what harvest rate would achieve this while accounting for natural population growth.

The target is to have the net rate of change (growth minus harvest) result in the desired decline. We can set up the equation:

$$ N_t = N_0 \times e^{(r - h) \times t} $$

Where $h$ is the harvest rate we want to solve for. Rearranging:

$$ h = r - \frac{\ln(N_t / N_0)}{t} $$

```{r}
# Starting population for harvest scenario
N_start <- 70000
N_target <- 50000
t_harvest <- 5

# Calculate required harvest rate
target_rate <- log(N_target / N_start) / t_harvest
harvest_rate <- r - target_rate
harvest_rate
```

The required annual harvest rate is **`r round(harvest_rate * 100, 2)`%** per year. This harvest rate offsets both the population growth and achieves the desired decline.

## Annual deer harvest numbers

Let's calculate how many deer need to be harvested each year to achieve this reduction:

```{r}
library(dplyr)
library(ggplot2)

# Simulation parameters
years <- 0:5
pop_size <- numeric(length(years))
harvest_numbers <- numeric(length(years))

# Initial conditions
pop_size[1] <- N_start

# Simulate population with harvest
for (i in 2:length(years)) {
  # Population grows
  pop_after_growth <- pop_size[i-1] * exp(r)
  # Calculate harvest number (proportion of post-growth population)
  harvest_numbers[i] <- pop_after_growth * harvest_rate
  # Population after harvest
  pop_size[i] <- pop_after_growth - harvest_numbers[i]
}

# Create data frame for results
harvest_df <- data.frame(
  Year = years[2:length(years)],
  Deer_Harvested = round(harvest_numbers[2:length(years)])
)

harvest_df
```

As we can see, the number of deer that need to be harvested decreases each year as the population declines.

## Visualizing the population trajectory

Let's create a visualization showing how the population changes over time with and without harvest:

```{r}
# Simulate population without harvest
pop_no_harvest <- N_start * exp(r * years)

# Create data frame for plotting
pop_df <- data.frame(
  Year = rep(years, 2),
  Population = c(pop_size, pop_no_harvest),
  Scenario = rep(c("With Harvest", "Without Harvest"), 
                 each = length(years))
)

# Create plot
ggplot(pop_df, aes(x = Year, y = Population, 
                   color = Scenario, linetype = Scenario)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(labels = scales::comma, 
                    limits = c(40000, 90000)) +
  labs(
    title = "Deer Population Trajectory Over 5 Years",
    subtitle = "Comparing growth scenarios with and without harvest management",
    x = "Year",
    y = "Population Size",
    color = "Scenario",
    linetype = "Scenario"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    plot.subtitle = element_text(size = 11),
    legend.position = "bottom"
  )
```

The plot clearly shows the diverging trajectories. Without harvest, the population would continue to grow exponentially, reaching over 85,000 individuals by year 5. With the calculated harvest rate, the population is reduced back to the target level of 50,000.

## Discussion and implications

The analysis reveals several important considerations for deer management:

1. **Substantial harvest required**: An annual harvest of around 13.5% is quite substantial. In the first year alone, nearly 9,000 deer would need to be harvested.

2. **Declining harvest numbers**: As the population decreases, so does the absolute number of deer that need to be harvested each year (from ~8,800 in year 1 to ~6,700 in year 5). This makes the management program progressively easier to implement.

3. **Continuous monitoring**: This model assumes the natural growth rate remains constant at 6.73% per year. In reality, growth rates can vary due to environmental conditions, disease, and density-dependent effects. Regular population monitoring would be essential.

4. **Alternative strategies**: The constant proportional harvest rate is just one approach. Managers might consider:
   - Constant absolute harvest numbers
   - Delayed harvest to assess population trends
   - Density-dependent harvest strategies that adjust based on population size
   - Spatial targeting of harvest in areas with highest densities

5. **Balancing interests**: The ~13% annual harvest rate required represents a significant management effort that must balance conservation goals (reducing impacts on native ecosystems) with sustainable harvest opportunities for hunters.

The mathematics demonstrate that returning to historical population levels is achievable but requires sustained, substantial harvest effort. This quantitative framework can inform policy discussions and help stakeholders understand the scale of management intervention needed.

## Code availability

All code for this analysis is embedded in this blog post and can be extracted from the source. The calculations are straightforward and can be easily adapted for other populations or management scenarios.
